---
title: "Exploring the heterogeneity in liver cancer scRNA-seq data with LDA"
author: "Xiyu Peng"
date: "Jan 2023"
output:
    html_document:
      toc: true
      toc_float: true 
      theme: united
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir= normalizePath('..'))
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


Here we walk through the pipeline of the latent Dirichlet Allocation (LDA) model, showing an example of its application on a scRNA-seq dataset from liver cancer patients.
This dataset is obtained from a recent published nature [paper](https://www.nature.com/articles/s41586-022-05400-x). 


# Starting point

In text mining, LDA requires the input as a document-by-term count matrix, where each row represent each document,each column represent each term, 
each entry in the matrix is the number of occurrences of each term (a word is a single occurrence of a term). 
Motivated by the similarities between text data mining and single-cell analysis, for single-cell analysis, LDA consider:

- **cells as words,**

- **cell types as terms,**

- **patient samples as documents,**

- **biological processes as topics.**


Before applying the LDA model, we also need to prepare the cell type count matrix as the input of LDA. 
One common approach to get the cell type count matrix is to pool all cells together and do the clustering.
For example, you can use *Seurat* to cluster cells and prepare the cell type-by-sample matrix. 
Before the pooled clustering analysis, we highly recommend to check if there is a batch effect across patients.

In this workflow, we use a published dataset to illustrate the application of LDA on single-cell dataset. 
We just downloaded the clustered data, including annotated cell types and cell type-by-sample count matrix from the paper. 
All data used in the analysis can be found in the supplementary data files of the original paper or in the [data portal](http://zhangninglab.com:3838/scPLC/).
All cell types have been well-annotated in the original paper.  


# Getting ready

Firstly, we have to install core R packages required for LDA analysis.
We use the R package *topicmodels* for model inference, R package *slam* for preparing the input data, R package *tidytext* for extracting the output data. 

```{r}
library(topicmodels)
library(slam)
library(tidytext)
```

In this workflow, we use some R packages for data visualization and analysis. But these R packages are not necesseray, you can use other similar R packages instead.

```{r}
library(tidyverse)
library(ComplexHeatmap)
library(survminer)
library(survival)
library(RColorBrewer)
```

Then we read the cell type count matrix, which was download from the the [data portal](http://zhangninglab.com:3838/scPLC/). We include 130 tumor tissue samples from 119 patients. Several patients have multiple tumor tissue samples.  

```{r}
metadata.sel.count_tumor_tissue <- read.csv("https://raw.githubusercontent.com/xiyupeng/LDA_examples/main/data/metadata-sel-count_tumor_tissue.csv",row.names = 1)

## exclude clusters that were removed in the paper due to low frequency or low quality
removed_clusters<-c("CD4T_01_CCR7", "CD4T_09_FOXP3", "CD8T_01_CCR7", "CD8T_02_CX3CR1", 
"CD8T_03_GZMK_S1PR1", "gdT_01_GNLY_S1PR5", "NK_01_FCGR3A_CX3CR1", 
"MonoDC", "Mo_01_CD14", "Mo_02_CD16", "Neu_02_S100A12", "Neu_03_ISG15", "Neu_04_TXNIP")

data_matrix<- metadata.sel.count_tumor_tissue %>% dplyr::select(!removed_clusters)
```

The final input data matrix for LDA has 130 samples with 76 features (clusters).

```{r}
dim(data_matrix)
```

# Model inference

We use *topicmodels* R package for model inference. The package provides two algorithms, variation EM and Gibbs sampling, for model inference. The two different algorithms may contribute to different results. 
See the [documentation](https://cran.microsoft.com/snapshot/2016-09-30/web/packages/topicmodels/vignettes/topicmodels.pdf) of the *topicmodels* R package.
But in the example given in the R package, the Gibbs sampling algorithm seems to return better models than Variation EM. Thus we prefer Gibbs sampling in the analysis. We used the following setting for Gibbs sampling: iter = 1000, burnin = 1000, thin = 100 (1000 Gibbs sampling draws are made with the first 1000 iterations discarded and then every 100th iteration kept).

Number of topics K also needs to be selected before running the model inference. Since the original paper identifies five TIME subtypes, we choose K = 5. You may use R package *ldatuning* or cross-validation to select number of topics K. 

The Gibbs sampling will take several minutes. We can save the result here for further analysis.

```{r,eval=FALSE}
## prepare data format
count_matrix_triplet <- as.simple_triplet_matrix(as.matrix(data_matrix))

## model inference 
lda<-LDA(count_matrix_triplet, control = list(seed = 2021, burnin = 1000,thin = 100, iter = 1000), k =5, method = "Gibbs")

## save the data
save(lda, file = "~/flow_cytometry/manuscript/liver cancer data/LDA_K5_filtered.Rdata")
```

Below we reload the data and check the output of the LDA model.

```{r}
## load the data
load(file = "~/flow_cytometry/manuscript/liver cancer data/LDA_K5_filtered.Rdata")
ntopics = 5

## extract output of the LDA model
para<-posterior(lda)
colnames(para$topic)<-paste0("topic",colnames(para$topic))
rownames(para$terms)<-paste0("topic",rownames(para$terms))
```

The main output of the LDA model contains two matrix: 

1. **Topic content matrix** (cell type-by-topic matrix) represents topics as different discrete distributions over cell types


```{r}
## show first five columns
head(round(para$terms[,1:5],5))
```

Or you can find clusters with top weight for each topic.

```{r}
terms(lda,5)
```

2. **Topic prevalence matrix** (topic-by-sample matrix) displays topic proportions estimated per sample


```{r}
## show first five rows
head(round(para$topics,3))
```

The topic prevalence matrix is merged with the original data frame for further analysis. But before we investigating the distribution of topics, we need to exam the content of each topic by selecting representative cell types.

```{r}
metadata.sel.count_tumor_tissue<-as.data.frame(cbind(metadata.sel.count_tumor_tissue,para$topic))
metadata.sel.count_tumor_tissue$pt<-gsub("_.*","",rownames(metadata.sel.count_tumor_tissue))
```

# Representative cell types


# Association to meta data

Our next step is to exam the prevalence of the topics and correlate to clinical outcomes.

Same as the paper, we only focus on tumor tissue samples from primary liver cancer patients with the three major histological subtypes (HCC, ICC, and CHC). The original paper identified five different TIME in tumor tissue samples, including immune activation (IA), immune suppression mediated by myeloid (ISM) or stromal cells (ISS), immune exclusion (IE), and immune residence (IR) phenotypes.

```{r,echo=FALSE,eval=FALSE}
## prepare meta data
clinical_data <- read.csv("~/flow_cytometry/manuscript/liver cancer data/liver_cancer_patient_clinical_data.csv",skip = 2)
clinical_data <- clinical_data[1:124,]
TIME_subtype<-read.csv("~/flow_cytometry/manuscript/liver cancer data/TIME_subtypes.csv",skip = 2)
clinical_data_merge<-merge(clinical_data,TIME_subtype,by = "Patient")
clinical_data_merge <- clinical_data_merge %>% select_if(~ !any(is.na(.)))
save(clinical_data_merge,file = "~/flow_cytometry/manuscript/liver cancer data/clinical_data.Rdata")
```
The first step is to prepare the meta data file, combining clinical data and assigned TIME subtypes as a single data frame. 
The clinical data file `liver_cancer_patient_clinical_data.csv` and TIME subtype data file `TIME_subtypes.csv` are supplementary data files from the original paper. The two data files can also be downloaded from [here](https://github.com/xiyupeng/LDA_examples/tree/main/data).


We have prepared the data frame and we just uploaded the pre-cleaned data. The original data files are the supplementary data files in the original paper.
The cleaned data file `clinical_data.Rdata` can be download from [here](https://github.com/xiyupeng/LDA_examples/raw/main/data/clinical_data.Rdata).

This data frame has 30 columns from 111 patients.
We first check column names of the 30 columns.

```{r}
load(file = "~/flow_cytometry/manuscript/liver cancer data/clinical_data.Rdata")
dim(clinical_data_merge)
colnames(clinical_data_merge)
```

We merge the clinical data `clinical_data_merge` with the scRNA-seq data matrix `metadata.sel.count_tumor_tissue`. For patient with multiple samples, only the first sample of the patient in the data matrix `metadata.sel.count_tumor_tissue` will be kept.

```{r}
data_merge<-merge(clinical_data_merge,metadata.sel.count_tumor_tissue,by.x = "Patient",by.y = "pt")
```

A heatmap can help to visualize the distribution of topics.

```{r,fig.height=12,fig.width=12}
## prepare data for the heatmap
pt_meta  <- data_merge %>%  select(Patient,Cancer_type_short,Differentiation,"Relapse_state..Yes.0.", "FPS_time","OS_state..Yes.0.","OS_time", TIMELASER)
gamma_data<-data_merge %>% select(starts_with("topic"))

## pre-identify patient subgroups with hierarchical clustering
rows.tree<-hclust(dist(gamma_data))
group<-as.data.frame(cutree(rows.tree,5))
colnames(group)<-"Group"
group$Group<-paste0("group",group$Group)

## select columns to be shown in the heatmap
pt_meta<-as.data.frame(cbind(pt_meta,group))
pt_meta_select<-pt_meta %>% dplyr::select(Group,Cancer_type_short,TIMELASER)

## Add color
ann_colors = list(
  Cancer_type_short = c(HCC = "#7570B3", CHC = "#E7298A", ICC = "#66A61E"),
  Group = c(group1 = "#999999", group2 = "#E69F00", group3 = "#56B4E9", group4 = "#009E73", group5 = "#F0E442"),
  TIMELASER = c(`TIME-IA`="#1B9E77", `TIME-IE`=  "#D95F02", `TIME-IR` = "#7570B3", `TIME-ISM` = "#E7298A" , `TIME-ISS` = "#66A61E")
)

## heatmap
ComplexHeatmap::pheatmap(gamma_data,annotation_row =pt_meta_select,cutree_rows = 5,annotation_colors = ann_colors)
```

# Survival Analysis

Kaplan-Meier analysis of OS and PFS stratified by five patient subgroups.

```{r}
data_merge$OS_state..Yes.0.<-ifelse(data_merge$OS_state..Yes.0.>0, 0, 1)
data_merge$Relapse_state..Yes.0.<-ifelse(data_merge$Relapse_state..Yes.0.>0, 0, 1)

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
my.labs = levels(as.factor(pt_meta$Group))

os <- Surv(data_merge$OS_time, data_merge$OS_state..Yes.0.)
ggsurvplot(survfit(os~Group,data = pt_meta),pval = TRUE,palette = cbPalette,censor.shape="|", censor.size=4,legend.labs=my.labs)+ylab("Overall Survival Probability")+xlab("Time(mos)")


pfs<-Surv(data_merge$FPS_time, data_merge$Relapse_state..Yes.0.)
ggsurvplot(survfit(pfs ~ Group, data = pt_meta),pval=TRUE,palette = cbPalette,censor.shape="|", censor.size=4,  legend.labs=my.labs)+ylab("Progression Free Probability")+xlab("Time(mos)")
```

Kaplan-Meier analysis of OS and PFS stratified by five TIME subtypes.

```{r}
TIMEPalette<-brewer.pal(n = 5, name = 'Dark2')
my.labs = levels(as.factor(pt_meta$TIMELASER))

ggsurvplot(survfit(os ~ TIMELASER,data = pt_meta),pval = TRUE,palette = TIMEPalette,censor.shape="|", censor.size=4,  legend.labs=my.labs)+ylab("Overall Survival Probability")+xlab("Time(mos)")

ggsurvplot(survfit(pfs ~ TIMELASER, data = pt_meta),pval=TRUE,palette = TIMEPalette,censor.shape="|", censor.size=4,  legend.labs=my.labs)+ylab("Progression Free Probability")+xlab("Time(mos)")
```


