---
title: "Analysis on a scRNA-seq data on liver tumor tissue"
output:
    html_document:
      toc: true
      toc_float: true 
      theme: united
---
Xiyu Peng

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir= normalizePath('..'))
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
library(corrplot)
library(tidyverse)
library(pheatmap)
library(topicmodels)
library(slam)
library(ggpubr)
library(tidytext)
library(survminer)
library(survival)
library(RColorBrewer)
```

* liver cancer data

```{r}
metadata.sel.count_tumor_tissue <- read.csv("~/flow_cytometry/manuscript/liver cancer data/metadata-sel-count_tumor_tissue.csv",row.names = 1)
removed_clusters<-c("CD4T_01_CCR7", "CD4T_09_FOXP3", "CD8T_01_CCR7", "CD8T_02_CX3CR1", 
"CD8T_03_GZMK_S1PR1", "gdT_01_GNLY_S1PR5", "NK_01_FCGR3A_CX3CR1", 
"MonoDC", "Mo_01_CD14", "Mo_02_CD16", "Neu_02_S100A12", "Neu_03_ISG15", "Neu_04_TXNIP")
data_matrix<-metadata.sel.count_tumor_tissue
data_matrix<- data_matrix %>% dplyr::select(!removed_clusters)
```

* heatmap

```{r}
M<-cor(data_matrix/rowSums(data_matrix))
ComplexHeatmap::pheatmap(M,heatmap_legend_param = list(title = "Correlation"))
```



```{r,eval=FALSE}
count_matrix_triplet <- as.simple_triplet_matrix(as.matrix(data_matrix))
lda<-LDA(count_matrix_triplet, control = list(seed = 2021, burnin = 1000,thin = 100, iter = 1000), k = 5, method = "Gibbs")
topics(lda)
terms(lda,10)
save(lda, file = "~/flow_cytometry/manuscript/liver cancer data/LDA_K5_filtered.Rdata")
ntopics<-5
```

# number of topics = 5

```{r}
ntopics<-5
load(file = "~/flow_cytometry/manuscript/liver cancer data/LDA_K5_filtered.Rdata")
para<-posterior(lda)
colnames(para$topic)<-paste0("topic",colnames(para$topic))
```


```{r,fig.width=5}
#### update to X50_res0.8
metadata.sel.count_tumor_tissue$topic<-topics(lda)
metadata.sel.count_tumor_tissue<-as.data.frame(cbind(metadata.sel.count_tumor_tissue,para$topic))

### beta
ap_topics <- tidy(lda, matrix = "beta")
ap_topics$topic<-as.factor(paste0("topic",ap_topics$topic))
#levels(ap_topics$topic)<-c("Activation topic", "Naive topic", "Exhaustion topic")

term_reorder<-as.character(1:21)
ggbarplot(ap_topics, x = "term", y = "beta",color = "topic",fill = "topic",facet.by = "topic",nrow = ntopics)+xlab("cell cluster")+ylab("topic weight")

ap_topics %>%
  group_by(topic) %>%
  slice_max(beta, n = 10) %>% 
  ungroup() %>%
  arrange(topic, -beta) %>%
  mutate(term = reorder_within(term, beta, topic)) %>%
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE,orientation = "y") +
  facet_wrap(~ topic, scales = "free") + 
  scale_y_reordered()+theme_bw()+
  xlab("topic weight")+ylab("cell cluster")
```


## select representative clusters with lift

```{r}
### calcualte the lift
wordcounts <- colSums(data_matrix)
logbeta <- lda@beta
emp.prob <- log(wordcounts) - log(sum(wordcounts))
lift <- logbeta - rep(emp.prob, each = nrow(logbeta))
colnames(lift)<-lda@terms
lift<-as.data.frame(lift)
lift$topic<-paste0("topic",1:ntopics)

### show top clusters for each topic
lift_top_terms <- lift %>% gather(.,key = cluster,value = lift,-topic) %>%
  group_by(topic) %>%
  slice_max(lift, n = 10) %>% 
  ungroup() %>%
  arrange(topic, -lift)

lift_top_terms %>%
  mutate(cluster = reorder_within(cluster, lift, topic)) %>%
  ggplot(aes(lift, cluster, fill = factor(topic))) +
  geom_col(show.legend = FALSE,orientation = "y") +
  facet_wrap(~ topic, scales = "free") + 
  scale_y_reordered()+theme_bw()+
  xlab("lift (log-ratio of topic weight over cell cluster frequency)")+ylab("cell cluster")
```


```{r}
### show top clusters for each topic
lift_all_terms <- lift %>% gather(.,key = cluster,value = lift,-topic) %>%
  group_by(topic) %>%
  ungroup() %>%
  arrange(topic, -lift)
```

## Link to TIME subtypes identified in the original paper

* TIME-IA

```{r,eval=FALSE}
cluster_list<- c("CD4T_07_CXCL13","CD4T_08_STMN1","CD8T_10_PDCD1","CD8T_13_STMN1","DC_03_LAMP3","Mph_06_CXCL9")

ap_topics %>% filter(term %in% cluster_list) %>% arrange(term,beta) %>% View()

ap_topics %>% filter(term %in% cluster_list) %>% filter(topic == "topic4")

lift_all_terms %>% filter(cluster %in% cluster_list) %>% arrange(cluster,lift) %>% View()

lift_all_terms %>% filter(cluster %in% cluster_list)  %>% filter(topic == "topic4")
```

* TIME-ISM

```{r,eval=FALSE}
cluster_list<- c("Mph_02_CCL20","Mph_03_SPP1","Neu_09_IFIT1","CD8T_13_STMN1","DC_03_LAMP3","Mph_06_CXCL9")

ap_topics %>% filter(term %in% cluster_list) %>% arrange(term,beta) %>% View()

ap_topics %>% filter(term %in% cluster_list) %>% filter(topic %in% c("topic2","topic4"))

lift_all_terms %>% filter(cluster %in% cluster_list) %>% arrange(cluster,lift) %>% View()

lift_all_terms %>% filter(cluster %in% cluster_list)  %>% filter(topic %in% c("topic2","topic4")) %>% arrange(cluster,lift) 
```

* TIME-ISS

```{r,eval=FALSE}
cluster_list<- c("CD4T_04_BAG3","CD4T_06_PLCG2","EC_03_TFF3","Fb_01_FAP","Mph_04_TREM2","NK_06_ITGA1")

ap_topics %>% filter(term %in% cluster_list) %>% arrange(term,beta) %>% View()

ap_topics %>% filter(term %in% cluster_list) %>% filter(topic %in% c("topic5")) %>% arrange(term,beta) 

lift_all_terms %>% filter(cluster %in% cluster_list) %>% arrange(cluster,lift) %>% View()

lift_all_terms %>% filter(cluster %in% cluster_list)  %>% filter(topic %in% c("topic5")) %>% arrange(cluster,lift)
```

* TIME-IE

```{r,eval=FALSE}
cluster_list<- c("CD8T_08_GZMK","EC_07_KDR_APOA2","EC_08_IGFBP3","Fb_01_FAP","Mph_04_TREM2","NK_06_ITGA1")

ap_topics %>% filter(term %in% cluster_list) %>% arrange(term,beta) %>% View()

ap_topics %>% filter(term %in% cluster_list) %>% filter(topic %in% c("topic3")) %>% arrange(term,beta) 

lift_all_terms %>% filter(cluster %in% cluster_list) %>% arrange(cluster,lift) %>% View()

lift_all_terms %>% filter(cluster %in% cluster_list)  %>% filter(topic %in% c("topic3")) %>% arrange(cluster,lift)
```

* TIME-IR

```{r,eval=FALSE}
cluster_list<- c("CD4T_03_GPR183","DC_01_CLEC9A","EC_01_CLEC4A","Mph_01_MARCO","Neu_07_APOA2","NK_05_CD160")

ap_topics %>% filter(term %in% cluster_list) %>% arrange(term,beta) %>% View()

ap_topics %>% filter(term %in% cluster_list) %>% filter(topic %in% c("topic3","topic1")) %>% arrange(term,beta) 

lift_all_terms %>% filter(cluster %in% cluster_list) %>% arrange(cluster,lift) %>% View()

lift_all_terms %>% filter(cluster %in% cluster_list)  %>% filter(topic %in% c("topic3","topic1")) %>% arrange(cluster,lift)
```

## link to patient level data

* Extract Gamma 

```{r}
gamma_data<-as.data.frame(para$topics)
colnames(gamma_data)<-paste0("topic",1:5)
```

* Incorporate Metadata

```{r}
metadata.sel.count_tumor_tissue$pt<- gsub("_.*","",rownames(gamma_data))

clinical_data <- read.csv("~/flow_cytometry/manuscript/liver cancer data/liver_cancer_patient_clinical_data.csv",skip = 2)
TIME_subtype<-read.csv("~/flow_cytometry/manuscript/liver cancer data/TIME_subtypes.csv",skip = 2)
metadata.sel.count_tumor_tissue$sample<-rownames(metadata.sel.count_tumor_tissue)
clinical_data_merge<-merge(metadata.sel.count_tumor_tissue,clinical_data,by.x = "pt", by.y = "Patient")
clinical_data_merge<-merge(clinical_data_merge,TIME_subtype,by.x = "pt", by.y = "Patient")
dim(clinical_data_merge)

#pt_meta<- clinical_data_merge %>% select(pt,Cancer_type_short,Differentiation,"Relapse_state..Yes.0.","FPS_time","OS_state..Yes.0.","OS_time", TIMELASER)
```

* Only use HCC, CHC, ICC subtypes

```{r,fig.height= 12, fig.width=12}
## only use HCC, CHC, ICC subtype
clinical_data_merge_sub <- clinical_data_merge %>% filter(Cancer_type_short %in% c("HCC","ICC","CHC")) 
pt_meta_sub  <- clinical_data_merge_sub %>% select(pt,Cancer_type_short,Differentiation,"Relapse_state..Yes.0.", "FPS_time","OS_state..Yes.0.","OS_time", TIMELASER)

gamma_data_sub<-gamma_data[clinical_data_merge_sub$sample,]
rows.tree<-hclust(dist(gamma_data_sub))

group<-as.data.frame(cutree(rows.tree,5))
colnames(group)<-"Group"
group$Group<-paste0("group",group$Group)

pt_meta_sub<-as.data.frame(cbind(pt_meta_sub,group))

pt_meta_select<-pt_meta_sub %>% dplyr::select(Group,Cancer_type_short,TIMELASER)

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
TIMEPalette<-brewer.pal(n = 5, name = 'Dark2')  #  "#1B9E77" "#D95F02" "#7570B3" "#E7298A" "#66A61E"

ann_colors = list(
  Cancer_type_short = c(HCC = "#7570B3", SLC = "darkgray", CHC = "#E7298A", ICC = "#66A61E", ASC = "#D95F02"),
  Group = c(group1 = "#999999", group2 = "#E69F00", group3 = "#56B4E9", group4 = "#009E73", group5 = "#F0E442", group6 = "#0072B2", group7 = "#D55E00", group8 = "#CC79A7"),
  TIMELASER = c(`TIME-IA`="#1B9E77", `TIME-IE`=  "#D95F02", `TIME-IR` = "#7570B3", `TIME-ISM` = "#E7298A" , `TIME-ISS` = "#66A61E")
)


ComplexHeatmap::pheatmap(gamma_data_sub,annotation_row =pt_meta_select,cutree_rows = 5,annotation_colors = ann_colors)

```


## survival analysis

* our subgroups

```{r}
table(pt_meta_sub$Group)
clinical_data_merge_sub$OS_state..Yes.0.<-ifelse(clinical_data_merge_sub$OS_state..Yes.0.>0, 0, 1)
clinical_data_merge_sub$Relapse_state..Yes.0.<-ifelse(clinical_data_merge_sub$Relapse_state..Yes.0.>0, 0, 1)

os <- Surv(clinical_data_merge_sub$OS_time, clinical_data_merge_sub$OS_state..Yes.0.)
ggsurvplot(survfit(os~Group,data = pt_meta_sub),pval = TRUE,palette = cbPalette)+ylab("Overall Survival Probability")+xlab("Month")


pfs<-Surv(clinical_data_merge_sub$FPS_time, clinical_data_merge_sub$Relapse_state..Yes.0.)
ggsurvplot(survfit(pfs ~ Group, data = pt_meta_sub),pval=TRUE,palette = cbPalette)+ylab("Progression Free Probability")+xlab("Month")
```

* TIMELASER

```{r}
ggsurvplot(survfit(os ~ TIMELASER,data = pt_meta_sub),pval = TRUE,palette = TIMEPalette)+ylab("Overall Survival Probability")+xlab("Month")

ggsurvplot(survfit(pfs ~ TIMELASER, data = pt_meta_sub),pval=TRUE,palette = TIMEPalette)+ylab("Progression Free Probability")+xlab("Month")
```

## Univariate cox regression

* Overall Survival

```{r}
## topic 1
os.cox <- coxph(os ~ topic1, data = clinical_data_merge_sub)
summary(os.cox)

## topic 2
os.cox <- coxph(os ~ topic2, data = clinical_data_merge_sub)
summary(os.cox)

## topic 3
os.cox <- coxph(os ~ topic3, data = clinical_data_merge_sub)
summary(os.cox)

## topic 4
os.cox <- coxph(os ~ topic4, data = clinical_data_merge_sub)
summary(os.cox)

## topic 4
os.cox <- coxph(os ~ topic5, data = clinical_data_merge_sub)
summary(os.cox)
```

* Progression-free Survival


```{r}
## topic 1
pfs.cox <- coxph(pfs ~ topic1, data = clinical_data_merge_sub)
summary(pfs.cox)

## topic 2
pfs.cox <- coxph(pfs ~ topic2, data = clinical_data_merge_sub)
summary(pfs.cox)

## topic 3
pfs.cox <- coxph(pfs ~ topic3, data = clinical_data_merge_sub)
summary(pfs.cox)

## topic 4
pfs.cox <- coxph(pfs ~ topic4, data = clinical_data_merge_sub)
summary(pfs.cox)

## topic 5
pfs.cox <- coxph(pfs ~ topic5, data = clinical_data_merge_sub)
summary(pfs.cox)
```
